---
title: "Improving LADD Coffee's Digital Strategy"
subtitle: "Proposal Presentation"
author: "Mickyas Shawel"
institute: "MSDM | Cal Poly Pomona"
date: today
format:
  revealjs:
    theme: default
    width: 1600
    height: 900
    footer: "Improving LADD Coffee's Digital Strategy"
    transition: fade
    transition-speed: fast
    incremental: false
    toc: true
    toc-depth: 1
    slide-number: true
    scrollable: true
    smaller: true
    code-fold: false
    code-overflow: wrap
    number-sections: false
    embed-resources: true
    css: styles.css
editor: visual
execute:
  echo: false
  warning: false
  message: false
  freeze: auto
---

```{r}
#| label: setup
#| include: false

library(tidyverse)
library(scales)

# Analyst-defined framework for the proposal.
# These are NOT observed LADD Coffee performance metrics.
channel_plan <- tribble(
  ~channel,                ~priority_score, ~primary_role,
  "Google Business/Search", 5,              "Capture local intent",
  "Instagram / Reels",      5,              "Build visual discovery",
  "TikTok / Short Video",   4,              "Create local awareness",
  "Website / Landing Page", 4,              "Own brand + conversion",
  "Email / SMS",            3,              "Drive repeat visits"
)

measurement_plan <- tribble(
  ~journey_stage, ~example_kpi, ~recommended_source,
  "Discover",     "Reach / profile views / search impressions", "Social insights + Google",
  "Consider",     "Menu views / direction requests / saves",    "Website + business profile",
  "Visit",        "Offer redemptions / attributed visits",      "POS + campaign codes",
  "Return",       "Repeat purchase / subscriber growth",        "POS + email/SMS"
)
```

# LADD Coffee: A New Local Brand

:::::: columns
::: {.column width="58%"}
## The opportunity

LADD Coffee is a newly opened specialty café in **West Covina, California**, operating at **987 S. Glendora Ave., Unit 3**.

Public coverage describes the shop as a **Korean specialty café** with visually distinctive drinks and desserts, including cream-top beverages, matcha drinks, and pastries.

For a new café, the digital challenge is not simply *being online*.

> The goal is to turn local digital discovery into a first visit — and a first visit into repeat behavior.
:::

:::: {.column width="42%"}
### Project question

**How can LADD Coffee build a measurable digital strategy that increases local discovery, store visits, and repeat engagement?**

::: {.callout-note title="Project status"}
This presentation is structured as an analytics and digital-strategy **proposal**. Proposed KPIs and priority scores are planning recommendations, not historical performance results.
:::
::::
::::::

::: notes
Open by emphasizing that LADD Coffee is a new business, so the strategy should prioritize discoverability and measurement infrastructure before complex optimization.
:::

# Table of Contents

1.  Business problem
2.  Analytics objective
3.  Customer and theory
4.  Data readiness
5.  Digital strategy
6.  Measurement plan
7.  Methods and next steps
8.  References

# Business Problem

## New shop, limited digital history

:::::: columns
:::: {.column width="52%"}
### Business challenge

A newly opened café has to solve three problems quickly:

::: incremental
- Become **discoverable** when nearby consumers search for coffee.
- Give customers a reason to choose LADD over established alternatives.
- Build enough first-party measurement to understand what actually drives visits and repeat purchases.
:::
::::

::: {.column width="48%"}
### Why this matters

Coffee decisions are often:

- **Local**
- **Mobile**
- **Visual**
- **Time-sensitive**
- Influenced by reviews, maps, social content, and peer recommendations

A fragmented digital presence can create awareness without producing measurable customer action.
:::
::::::

# Analytics Objective

## One primary objective

> **Determine which digital touchpoints most effectively move nearby consumers from awareness to an in-store visit, then use that evidence to prioritize LADD Coffee's marketing effort.**

### Supporting questions {.incremental}

- Which channels generate the most qualified local discovery?
- Which menu items create the strongest visual/content hooks?
- Which digital actions are most associated with store-intent signals?
- Which customers can be encouraged to return?

# Importance

## Build the measurement system early

For an established retailer, analytics often means optimizing years of historical data.

For LADD Coffee, the immediate advantage is different:

::: incremental
1.  Define useful KPIs **before** large amounts of data accumulate.
2.  Establish consistent campaign tagging and channel attribution.
3.  Connect discovery metrics with store-intent and purchase signals.
4.  Learn which creative and products generate repeatable demand.
:::

::: {.callout-tip title="Strategic advantage"}
A new business can design cleaner measurement habits now instead of trying to repair fragmented data later.
:::

# Customer Behavior Framework

## A journey-based approach

The strategy can be organized around a simple customer journey:

**Discover → Consider → Visit → Return**

::::: columns
::: {.column width="50%"}
### Discover & Consider

Potential customers encounter:

- Google Maps / local search
- Instagram and short-form video
- Creator or customer posts
- Reviews and recommendations
- Menu/product imagery
:::

::: {.column width="50%"}
### Visit & Return

The digital experience should make it easy to:

- Get directions
- Understand the menu
- Identify signature items
- Save/share the shop
- Join an owned audience
- Receive a reason to return
:::
:::::

::: notes
Connect this to consumer decision journeys: digital media should reduce uncertainty and make the next action obvious.
:::

# Current Public Signals

## What is already visible online

Public sources provide several useful early signals:

- LADD Coffee LLC was filed in California in **March 2026**.
- A June 2026 local feature described LADD as a newly opened Korean specialty café in West Covina.
- Public descriptions emphasize **cream-top drinks, matcha, desserts, and a minimalist café environment**.
- The physical location is listed as **987 S Glendora Ave., Unit 3, West Covina**.

These signals are useful for positioning, but they are **not enough to evaluate marketing performance**.

::: {.callout-warning title="Data limitation"}
Public listings can describe the business, but they cannot answer which channels drive store visits, revenue, repeat customers, or customer lifetime value.
:::

# Data Adequacy

## What we have vs. what we need

```{r}
#| label: tbl-measurement
#| tbl-cap: "Recommended measurement structure for the LADD Coffee digital customer journey"

measurement_plan |>
  knitr::kable(
    col.names = c("Journey Stage", "Example KPI", "Recommended Data Source"),
    align = c("l", "l", "l")
  )
```

The key gap is **first-party behavioral and transaction data**.

# Additional Data Needed

## Build a practical data foundation

::: incremental
- Google Business Profile performance
- Website or landing-page analytics
- UTM-tagged campaign traffic
- Instagram/TikTok content metrics
- POS transactions by day/time/product
- Promotion or QR-code redemptions
- Email/SMS opt-ins
- Repeat-customer indicators
:::

### Recommended unit of analysis

Where privacy and platform limitations allow, aggregate performance by:

**Date × Channel × Campaign/Content × Customer Action**

This structure supports meaningful comparisons without requiring personally identifiable customer data.

# Digital Strategy

## Prioritize channels by customer role

```{r}
#| label: fig-channel-priority
#| fig-cap: "Proposed channel priorities for LADD Coffee's launch-stage digital strategy"
#| fig-width: 9
#| fig-height: 5.5

channel_plan |>
  mutate(channel = fct_reorder(channel, priority_score)) |>
  ggplot(aes(x = priority_score, y = channel)) +
  geom_col() +
  geom_text(
    aes(label = primary_role),
    hjust = -0.05,
    size = 4.2
  ) +
  scale_x_continuous(
    limits = c(0, 7.5),
    breaks = 0:5
  ) +
  labs(
    x = "Proposed Priority Score (1–5)",
    y = NULL,
    caption = "Analyst-defined planning scores; not observed performance data."
  ) +
  theme_minimal(base_size = 15) +
  theme(
    panel.grid.major.y = element_blank(),
    plot.caption = element_text(face = "italic")
  )
```

::: notes
Explain that this is a reproducible chart generated from a small planning dataset in R. The scores are recommendations, not measured results.
:::

# Channel Roles

## Google: capture intent

Optimize the business profile for:

- Accurate hours and location
- Menu/product photos
- Signature-item language
- Review generation and response
- Direction and call actions

## Social: manufacture discovery

Use Instagram and TikTok to make the product visually memorable:

::: incremental
- Cream-top preparation
- Matcha and specialty drink builds
- Dessert pairings
- "What to order first" content
- Customer reactions and UGC
- West Covina / SGV local discovery hooks
:::

# Owned Digital Experience

## Give discovery somewhere to convert

A lightweight website or landing page should answer the questions a first-time customer has immediately:

::::: columns
::: {.column width="50%"}
### Essential content

- Location and hours
- Menu / signature products
- High-quality photography
- Directions
- Social links
- FAQ
:::

::: {.column width="50%"}
### Measurement

- GA4
- UTM conventions
- Direction-click events
- Menu-view events
- Email/SMS signup
- Offer/QR redemptions
:::
:::::

The website should function as a **conversion and measurement layer**, not merely an online brochure.

# Content Strategy

## Make signature products the creative engine

Instead of generic "come visit us" posts, content should give audiences something specific to remember.

### Content pillars {.incremental}

1.  **Signature drinks** — cream tops, matcha, seasonal specialties
2.  **Craft/process** — preparation, ingredients, barista expertise
3.  **Dessert pairing** — drinks + pastries
4.  **Local identity** — West Covina and surrounding community
5.  **Social proof** — reviews, reactions, UGC, creator visits

::: aside
Public descriptions of LADD already emphasize visually distinctive drinks and desserts, making product-led storytelling a logical starting point.
:::

# Measurement Framework

## KPIs by stage

| Stage | Primary Question | Example KPIs |
|------------------------|------------------------|------------------------|
| Discover | Are nearby consumers finding LADD? | Search impressions, reach, profile views, video views |
| Consider | Are they showing intent? | Menu views, saves, shares, direction clicks |
| Visit | Are digital efforts producing action? | Redemptions, attributed transactions, store traffic proxies |
| Return | Are we building customer value? | Repeat purchase, subscriber growth, offer response |

# Methods

## Proposed analytics workflow

::: incremental
1.  **Instrument** the customer journey with consistent events and campaign tags.
2.  **Collect** channel, content, POS, and promotion data.
3.  **Wrangle** data into a common date/channel/campaign structure.
4.  **Describe** performance with funnel and cohort metrics.
5.  **Compare** channels and creative using normalized KPIs.
6.  **Test** offers, products, and creative systematically.
7.  **Recommend** budget and content changes from observed results.
:::

# Data Wrangling

## Example structure in R

```{r}
#| echo: true

campaign_data <- tibble(
  date = as.Date(c("2026-08-01", "2026-08-01", "2026-08-02")),
  channel = c("Instagram", "Google", "TikTok"),
  impressions = c(4200, 1800, 5100),
  actions = c(210, 175, 255)
) |>
  mutate(
    action_rate = actions / impressions,
    channel_group = case_when(
      channel == "Google" ~ "Intent",
      TRUE ~ "Discovery"
    )
  )

campaign_data
```

::: callout-note
The values above are **illustrative practice data** showing how future LADD Coffee campaign data could be structured and wrangled. They are not actual business results.
:::

# Sample Characteristics

## What a useful future sample should contain

A meaningful evaluation period should capture:

- Weekdays and weekends
- Multiple dayparts
- Several content formats
- Paid and organic traffic
- More than one promotion or creative treatment
- Product-level purchase variation
- Enough time to observe repeat behavior

The first goal is not a huge dataset. It is a **consistent dataset**.

# Measures

## Core definitions

**Action Rate**

\[ \text{Action Rate} = \frac{\text{Desired Digital Actions}} {\text{Impressions or Sessions}} \]

**Offer Redemption Rate**

\[ \text{Redemption Rate} = \frac{\text{Redeemed Offers}} {\text{Distributed or Viewed Offers}} \]

**Repeat Purchase Rate**

\[ \text{Repeat Purchase Rate} = \frac{\text{Returning Customers}} {\text{Customers Observed}} \]

# Analytics Methods

## Start simple, then increase sophistication

### Phase 1 — Descriptive

- Channel performance
- Content performance
- Product mix
- Day/time patterns
- Funnel conversion

### Phase 2 — Diagnostic

- Campaign comparisons
- Cohort behavior
- Promotion lift
- Creative-level differences

### Phase 3 — Predictive

Once sufficient data exist:

- Visit/purchase propensity
- Repeat-purchase classification
- Customer or product segmentation
- Demand forecasting

# Recommendations

## First 90 days

::: incremental
### 1. Own local search

Treat Google Business Profile and local-search accuracy as core conversion infrastructure.

### 2. Build product-led social content

Use signature beverages and desserts as the center of short-form creative.

### 3. Establish first-party measurement

Launch a simple owned landing experience with GA4, UTMs, event tracking, and opt-in capture.

### 4. Connect marketing to transactions

Use trackable offers, QR codes, POS tags, or campaign-specific promotions to create attribution signals.

### 5. Create a test-and-learn rhythm

Review performance monthly and move effort toward channels and creative that produce meaningful customer actions.
:::

# Expected Value

## From visibility to learning

A stronger digital strategy should help LADD Coffee:

- Increase qualified local awareness
- Make signature products easier to discover
- Reduce friction between discovery and visit
- Create an owned audience for retention
- Learn which channels and content influence behavior
- Make future marketing investment more evidence-based

> The most valuable early outcome is not simply more impressions — it is a system that tells LADD Coffee what is working.

# References

- CaliNook. (2026). *LADD Coffee*. June 3, 2026.
- Los Angeles County fictitious business name publication. (2026). *LADD Coffee*, West Covina.
- Corner. (2026). *LADD Coffee — West Covina*.
- Quarto. *RevealJS Presentations*.
- CCIDM. *Effective Quarto Slides*.

# Appendix

## Research notes

This proposal uses publicly available information to establish basic business context. No private sales, POS, web analytics, advertising, or customer-level data from LADD Coffee were available during preparation.

## Reproducibility

The channel-priority chart and planning tables in this deck are generated directly from R objects contained in the QMD file.

## Future project links

**Published presentation:** <https://mjshawell.github.io/IBM-6540-IBM-6300/Improving-LADD-Coffee-Digital-Strategy.html#/title-slide>

**GitHub repository:** <https://github.com/mjshawell/IBM-6540-IBM-6300>
